---
title: "Administrative Boundaries"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

#### Data set details

|||
| ----------- | ----------- |
| **Data set description:** | Administrative boundaries for different countries |
| **Source:**   | [geoBoundaries](https://www.geoboundaries.org/) |
| **Details on the retrieved data:**   | Data on different levels of administrative boundaries for countries such as Nigeria and Chad. |
| **Spatial and temporal resolution:**  | World administrative boundaries. |

# Downloading administrative boundaries of countries with `rgeoboundaries`

The [`rgeoboundaries`](https://github.com/wmgeolab/rgeoboundaries) package is a client for the [geoBoundaries API](https://www.geoboundaries.org/), providing country political administrative boundaries.
This tutorial takes you through the steps of downloading administrative boundaries of countries using the `rgeoboundaries` package and visualising them using the `ggplot2` package.


## Installing `rgeoboundaries`

The `rgeoboundaries` package can be downloaded from GitHub using the `remotes` package which allows easy installation of R packages from remote repositories such as GitHub.
We install and load the `remotes` package and use it to install the `rgeoboundaries` package from GitHub as follows.

```{r install-rgeoboundaries, eval=FALSE}
# install.packages("remotes")
library(remotes)
remotes::install_github("wmgeolab/rgeoboundaries")
```

## Administrative boundaries of a single country

To download boundaries of countries we use the `geoboundaries()` function of `rgeoboundaries`.
For example, we can download the administrative boundary of Nigeria and assign it to a variable called `nigeria_boundary` as follows.

```{r download-nigeria-boundary}
library(rgeoboundaries)
nigeria_boundary <- geoboundaries("Nigeria")
```

The `ggplot2` package can be used to plot the administrative boundaries downloaded.
`ggplot2` allows us to easily visualise simple feature objects using the `geom_sf()` function and can be used to plot the administrative boundary of Nigeria as follows.


```{r vis-nigeria-boundary}
# install.packages("ggplot2")
library(ggplot2)
ggplot(data = nigeria_boundary) +
  geom_sf()
```

## Administrative boundaries of multiple countries


We can also download the boundaries of multiple countries together by including the names of countries as a vector. See how the boundaries of Nigeria and Chad are downloaded below.

```{r download-nigeria-chad-boundary}
nigeria_chad_boundaries <- geoboundaries(c("Nigeria", "Chad"))
```

```{r}
ggplot(data = nigeria_chad_boundaries) +
  geom_sf()
```

## Different levels of administrative boundaries


If available, lower levels of administrative boundaries in countries can be downloaded too. We just have to pass the administrative level as an argument in the `geoboundaries()` function. Administrative level 1 ("adm1") is the highest level, while administrative level 5 ("adm5") is the lowest. This means the country will be further sub-divided into administrative divisions as the Administrative level progresses from 1 to 5.
See how the first and second administrative level boundaries of Nigeria and Chad are downloaded below.

```{r download-nigeria-chad-lowerlvl-boundary}
# downloading administrative level 1 boundaries
nigeria_chad_admlvl1_boundaries <- geoboundaries(c("Nigeria", "Chad"), "adm1")

ggplot(data = nigeria_chad_admlvl1_boundaries) +
  geom_sf()

# downloading administrative level 2 boundaries
nigeria_chad_admlvl2_boundaries <- geoboundaries(c("Nigeria", "Chad"), "adm2")

ggplot(data = nigeria_chad_admlvl2_boundaries) +
  geom_sf()
```

# Understanding the downloaded data

If we print out the downloaded administrative boundary of Nigeria, we will see that the downloaded boundary is a simple feature collection with 1 feature and 5 fields.
A feature is thought of as a single object, Nigeria in this case. The fields are thought of as the geometrical attributes included about each feature; `shapeName`, `shapeISO`, `shapeID`, `shapeGroup` and `shapeType` in the above object.

```{r print-nigeria-boundary}
nigeria_boundary
```

The simple feature object retrieved with the boundaries of Nigeria and Chad is a simple feature collection with 2 features and 5 fields.
The 2 features will refer to the two countries (Nigeria and Chad), and the features will be the geometrical attributes of each feature.

```{r print-nigeria-chad-boundary}
nigeria_chad_boundaries
```

Similarly, downloads of administrative level 1 (`nigeria_chad_admlvl1_boundaries`) and 2 (`nigeria_chad_admlvl2_boundaries`) boundaries of Nigeria and Chad will result in simple feature objects with 60 and 843 features respectively. Here each feature will be a single administrative division. The number of administrative divisions increasing as the countries are further sub-divided from administrative level 1 divisions to administrative level 2 divisions.


The `shapeName` field of the simple feature objects has the names of the administrative divisions.
We can create a map with the names of the divisions for Nigeria and Chad using the `geom_sf_label()` function and setting `label = shapeName`. 

```{r vis-nigeria-chad-boundary-wnames}
ggplot(data = nigeria_chad_boundaries) +
  geom_sf() +
  geom_sf_label(aes(label = shapeName))
```

Labelling the administrative divisions also automatically labelled the axes as `x` and `y`. If we want to change these axis labels, we can use `xlab()` and `ylab()` functions. We can also use the `ggtitle()` function to add a title to the plot.

```{r vis-nigeria-chad-wtitle}
ggplot(data = nigeria_chad_boundaries) +
  geom_sf() +
  geom_sf_label(aes(label = shapeName)) +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Nigeria and Chad boundaries")
```



# Examples

The following is an example on how to use the `rgeoboundaries` package to download administrative boundaries and `ggplot2` to visualise them.

```{r rgeoboundaries-example}
# remotes::install_github("wmgeolab/rgeoboundaries")
# install.packages("ggplot2")
library(rgeoboundaries)
library(ggplot2)

# Country boundaries of Nigeria and Chad
nigeria_chad_boundaries <- geoboundaries(c("Nigeria", "Chad"))
ggplot(data = nigeria_chad_boundaries) +
  geom_sf() +
  geom_sf_label(aes(label = shapeName)) +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Country boundaries")

# Administrative Level 1 boundaries of Nigeria and Chad
nigeria_chad_admlvl1_boundaries <- geoboundaries(c("Nigeria", "Chad"), "adm1")
ggplot(data = nigeria_chad_admlvl1_boundaries) +
  geom_sf() +
  ggtitle("Administrative Level 1 boundaries")
```

The next example uses the `rgeoboundaries` package to download administrative boundaries and `leaflet` to visualise them interactively.


```{r leaflet-example}
# remotes::install_github("wmgeolab/rgeoboundaries")
# install.packages("leaflet")
library(rgeoboundaries)
library(leaflet)

# Country boundaries of Nigeria and Chad
nigeria_chad_boundaries <- geoboundaries(c("Nigeria", "Chad"))

nigeria_chad_boundaries %>%
  leaflet() %>%
  addTiles() %>%
  addPolygons(label = nigeria_chad_boundaries$shapeName)


# Administrative Level 1 boundaries of Nigeria and Chad
nigeria_chad_admlvl1_boundaries <- geoboundaries(c("Nigeria", "Chad"), "adm1")

nigeria_chad_admlvl1_boundaries %>%
  leaflet() %>%
  addTiles() %>%
  addPolygons(label = nigeria_chad_admlvl1_boundaries$shapeName)
```


# References

- `rgeoboundaries` repository: https://github.com/wmgeolab/rgeoboundaries  
- `ggplot2` package: https://ggplot2.tidyverse.org/


---

<p style ="color:grey; font-size: 13px;">  
Last updated: `r Sys.Date()`  
Source code: https://github.com/rspatialdata/rspatialdata.github.io/blob/main/admin_boundaries.Rmd 
<details>
<summary><span style ="color:grey; font-size: 13px;">Tutorial was complied using: (click to expand)</span></summary>
```{r echo=FALSE}
sessionInfo()
```
</details>
</p>
