---
title: "Malaria"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DT)
library(ggthemes)
```

#### Data set details

|||
| ----------- | ----------- |
| **Data set description:** | Malaria data in different locations |
| **Source:**   | [Malaria Atlas Project](https://malariaatlas.org/) |
| **Details on the retrieved data:**   | Parasite rate (PR) survey data and vector occurrence data in Zimbabwe and India. |
| **Spatial and temporal resolution:**  | Malaria data at country level (with information on specific sites). |

# Downloading and visualising malaria data with `malariaAtlas`

This tutorial shows how to use the `malariaAtlas` package to download, visualize and manipulate global malaria data hosted by the [Malaria Atlas Project](https://malariaatlas.org/). These data can be explored and downloaded interactively at [data explorer](https://malariaatlas.org/explorer/#/).

The `malariaAtlas` package enables users to download three types of data:

- parasite rate (PR) survey data (*Plasmodium falciparum* and *Plasmodium vivax*)
- vector occurrence data
- administrative boundary shapefiles to visualise data
- rasters covering a range of modelled outputs related to malaria research such as predicted malaria parasite prevalence

 
## Installation of `malariaAtlas` package

The `malariaAtlas` package can be installed from CRAN and loaded as follows:

```{r install-malariaAtlas, eval=FALSE}
install.packages("malariaAtlas")
```

```{r load-lib}
library(malariaAtlas)
```


## Checking available data

The `malariaAtlas` package provides users with a convenient API to interact with the publicly available malaria data hosted by the Malaria Atlas Project.
The `listData()` function helps retrieve a list of all the available data which can be downloaded, and is a wrapper for `listPoints()`, `listRaster()` and `listShp()`.
We can find all countries that have PR survey points data with `listData(datatype = "pr points")`,
and all countries with vector occurrence data with `listData(datatype = "vector points")`.
We can also see raster data available with `listData(datatype = "raster")`,
and shapefiles available with `listData(datatype = "shape")`.


The `isAvailable_pr()` function can be used to check whether a specified country or location has PR survey points.
Similarly, `isAvailable_vec()` checks whether a specified country or location has vector occurrence data.
For example, we can check whether Zimbabwe has PR survey points and India has vector occurrence data as follows:

```{r func_pr, message = FALSE}
isAvailable_pr(country = "Zimbabwe")
isAvailable_vec(country = "India")
```


## Dowloading and visualising data


### PR survey data

The `getPR()` function downloads all the publicly available PR points for a country (or countries) and returns it as a dataframe. The `species` argument is a string specifying the Plasmodium species and can be `Pf`, `Pv` or `BOTH`.
 
```{r get-pr,message = FALSE,warning=FALSE}
ZM_PR_data <- getPR(country = "Zimbabwe", species = "BOTH")
```

```{r, echo=FALSE}
ZM_PR_data %>%
  datatable(extensions = c("Scroller", "FixedColumns"), options = list(
    deferRender = TRUE,
    scrollY = 350,
    scrollX = 350,
    dom = "t",
    scroller = TRUE,
    fixedColumns = list(leftColumns = 4)
  ))
```

`autoplot()` can be used to quickly and easily visualise the downloaded PR survey points.
 
```{r viz-PR, message=FALSE,warning=FALSE}
autoplot(ZM_PR_data)
```


### Vector data

 
The `getVecOcc()` function downloads all publicly available vector points for a country (or countries) and returns it as a dataframe. The `species` argument is a string specifying the Anopheles species for finding the vector occurrences points, the options are `Anopheles....` or `ALL`.
 
```{r get-vec,message = FALSE,warning=FALSE}
IN_VEC_data <- getVecOcc(country = "India", species = "ALL")
```

```{r, echo=FALSE}
IN_VEC_data %>%
  datatable(extensions = c("Scroller", "FixedColumns"), options = list(
    deferRender = TRUE,
    scrollY = 350,
    scrollX = 350,
    dom = "t",
    scroller = TRUE,
    fixedColumns = list(leftColumns = 4)
  ))
```

`autoplot()` uses `ggplot2` to visualise vector occurrence data.

```{r message = FALSE}
autoplot(IN_VEC_data)
```


### Shapefiles

The `getShp()` function downloads a shapefile for a specified country (or countries).
The `as.MAPshp()` function converts the SpatialPolygon downloaded using `getShp()` into a `MAPshp` object (data.frame) for easy plotting with `autoplot()`.

```{r message = FALSE}
IN_shp <- getShp(ISO = "IND", admin_level = c("admin1"))
IN_shp <- as.MAPshp(IN_shp)
autoplot(IN_shp)
```

### Modelled Rasters

The `getRaster()` function downloads publicly available rasters for a specific surface and year, clipped to a given bounding box or shapefile. For example we can download Plasmodium falciparum PR2-10 for Zimbabwe in 2010 as follows:
 
```{r message = FALSE}
ZM_shp <- getShp(country = "Zimbabwe", admin_level = "admin0")
ZMB_Pf <- getRaster(surface = "Plasmodium falciparum PR2-10", shp = ZM_shp, year = 2010)
```
 
Note that to use downloaded rasters and shapefiles directly with `autoplot()`, they must first be converted to usable rasters and shapefiles data frames, hence we use the `as.MAPraster()` and `as.MAPShp()` respectively.


```{r message = FALSE}
ZMB_Pf <- as.MAPraster(ZMB_Pf)
ZMB_shp_data <- as.MAPshp(ZM_shp)
autoplot(ZMB_Pf, shp_df = ZMB_shp_data)
```

## Complete visualisation examples


Since the `ggplot2` is loaded as dependency when the `malariaAtlas` package is loaded, the functionalities of `ggplot2` can also be used for more appeal and options.


### PR survey points

```{r message = FALSE}
ZMB_shp <- getShp(country = "Zimbabwe", admin_level = "admin0") # get shapefile
ZMB_shp_df <- as.MAPshp(ZMB_shp) # convert shapefile
ZMB_PfPR2_10 <- getRaster(surface = "Plasmodium falciparum PR2-10", shp = ZMB_shp, year = 2013) # get raster
ZMB_PfPR2_10_df <- as.MAPraster(ZMB_PfPR2_10) # convert raster
pr <- getPR(country = c("Zimbabwe"), species = "Pf") # get PR data

p <- autoplot(ZMB_PfPR2_10_df, shp_df = ZMB_shp_df, printed = FALSE) # plot raster and shapefile
p[[1]] + geom_point(
  data = pr[pr$year_start == 2013, ],
  aes(longitude, latitude, fill = positive / examined, size = examined), shape = 21
) +
  scale_size_continuous(name = "Survey Size") +
  scale_fill_distiller(name = "PfPR", palette = "RdYlBu") +
  ggtitle("Raw PfPR Survey points\n + Plasmodium falciparum PR2-10 in Zimbabwe in 2013") +
  theme_solarized() +
  labs(x = "Longitude", y = "Latitude")
```

### Vector points

```{r message = FALSE}
IND_shp <- getShp(country = "India", admin_level = "admin0") # get shapefile
IND_shp_df <- as.MAPshp(IND_shp) # convert shapefile
IND_an_dirus <- getRaster(surface = "Anopheles dirus species complex", shp = IND_shp) # get raster
IND_an_dirus_df <- as.MAPraster(IND_an_dirus) # convert raster
vec <- getVecOcc(country = c("India"), species = "Anopheles dirus") # get vector point data

p <- autoplot(IND_an_dirus_df, shp_df = IND_shp_df, printed = FALSE) # plot raster and shapefile
p[[1]] + geom_point(data = vec, aes(longitude, latitude), shape = 21, show.legend = TRUE) +
  scale_fill_distiller(name = "Predicted distribution", palette = "YlOrRd") +
  ggtitle("Raw Vector Survey points + \nThe predicted distribution of Anohpeles dirus species complex") +
  theme(legend.position = "bottom") +
  theme_solarized() +
  labs(x = "Longitude", y = "Latitude")
```



## References

- The `malariaAtlas` package repository : https://github.com/malaria-atlas-project/malariaAtlas
- Malaria Atlas Project: https://malariaatlas.org/
- Malaria data explorer: https://malariaatlas.org/explorer/#/


---

<p style ="color:grey; font-size: 13px;">  
Last updated: `r Sys.Date()`  
Source code: https://github.com/rspatialdata/rspatialdata.github.io/blob/main/malaria.Rmd 
<details>
<summary><span style ="color:grey; font-size: 13px;">Tutorial was complied using: (click to expand)</span></summary>
```{r echo=FALSE}
sessionInfo()
```
</details>
</p>
